Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-08-18 Manuscript/manuscript.tex
Tab. -> Table

For a closed-shell reference determinant ($s_0=0$), hCI0 reduces to the restricted HF determinant.
The first non-trivial order is hCI1 (see left of Fig.~\ref{fig:hCI}), which accounts for all single excitations (as CIS)
plus all paired doubled excitations (two electrons promoted from the same occupied orbital into the same virtual orbital),
as shown in the top left of Fig.~\ref{fig:determinants}.
For both classes of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, hence the empty blocks at $s=0$ for odd $e$.
hCI1.5 augments the set of hCI1 determinants with the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 incorporates all classes of determinants where $\Ndet$ scale as $N^4$.
In total, it accounts for all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons,
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Likewise, in eCI, $\Ndet$ scales as $N^{2e}$.
The computational scaling along the two routes is shown in Tab.~\ref{tab:scaling}.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space specified by the reference determinant(s) and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the present eCI, sCI and FCI calculations.
In a given calculation, the energies are considered to be converged when the (largest) EN2 correction computed in the truncated Hilbert space
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delivering fairly converged energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we further evaluated the standard and the renormalized EN2 perturbative correction. \cite{Garniron_2019}
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for eCI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation.
The computational scaling associated with the CI+EN2 calculations is also presented in Tab.~\ref{tab:scaling}.

@@

We start the discussion with the excitations from closed-shell systems, with corresponding statistical errors shown in Tab.~\ref{tab:stat1}.
The first level of hCI, hCI1, produces poor excitation energies, with a MAE of \SI{1.07}{\eV}.
This model shares the same computational scaling with CIS, but includes the paired double excitations,
which clearly worsen the decent CIS results (MAE of \SI{0.61}{\eV}).
Moving one rank up, to hCI1.5, the MAE increases to \SI{1.91}{\eV}, and then reaches a maximum of \SI{3.53}{\eV} at the hCI2 level.
We notice that hCI2 delivers somewhat smaller errors than CISD (MAE of \SI{4.09}{\eV}), even though both are way too large.
From that point on, the next hCI models generate progressively better results.
Despite the improvement with respect to hCI2, hCI2.5 is still as inaccurate (MAE of \SI{1.95}{\eV}) as hCI1.5.
At the hCI3 level, significantly smaller errors are finally achieved (MAE of \SI{0.19}{\eV}).
This is close to that obtained with the eCI model of same order, CISDT (\SI{0.17}{\eV}).
Even smaller errors are produced at the hCI3.5 level (MAE of \SI{0.11}{\eV}), but at a considerable computational cost.

@@

We now shift to the discussion of excitations for open-shell systems.
The statistical errors are shown in Tab.~\ref{tab:stat2}.
The key difference on the calculations for open-shell excitations is that the same type of reference was employed for ground- and excited-state calculations, namely,
a single open-shell determinant, shown in the center panel of Fig.~\ref{fig:hCI}.
In contrast, the excited states from the closed-shell systems relied on different classes of reference determinants.
This accounts for the more accurate results observed for open-shell excitations and explains most of the differences with respect to excitations from closed-shell systems, as discussed in detail in the following.

@@

We refer back to Tab.~\ref{tab:stat1} to discuss the performance of sCI for excited states of closed-shell systems.
The first model, sCI2/sCI2, systematically overestimates the excitation energies, producing a MAE of \SI{1.34}{\eV}.
The same two factors discussed above for hCI explain such large errors.
First, the ground state description is favored due to the use of ground-state HF orbitals.
Indeed, with state-specific orbitals, the $\Delta$sCI2/sCI2 model reduces the MAE to \SI{0.79}{\eV}, though still overestimating the excitation energies.
Second, there is an unbalanced description of correlation for ground and excited states, associated with the classes of determinants.
While sCI2/sCI2 accounts for an additional $s=2$ sector for the ground state (which can be qualitatively described in the $s=0$ sector),
no addition pair of electrons is allowed to become unpaired in the excited state calculation, even though the state is qualitatively described by a determinant that is contained in the $s=2$ sector.
In other words, unpaired excitations are allowed to correlate the ground state, but not the excited state, given their respective closed-shell and open-shell characters,
thus creating a bias towards the former.

@@

The situation for open-shell systems is quite different, in close analogy to the previous discussion regarding hCI.
As shown in Tab.~\ref{tab:stat2}, sCI1 provides considerably more accurate excitation energies for open-shell than for closed-shell systems.
The reason should not be surprising at this point.
Both ground and excited states of open-shells can be qualitatively described by the same type of reference (a single $s=1$ open-shell determinant).
Moreover, the state-specific approach is superior.
sCI1 presents a MAE of \SI{0.39}{\eV} and overestimates the excitation energies (MSE of \SI{+0.24}{\eV}), in view of the bias introduced by the ground-state orbitals.
With state-specific orbitals, $\Delta$sCI1 reduces the MAE to \SI{0.27}{\eV} and the MSE to \SI{-0.11}{\eV}, which is smaller in absolute value than found for sCI1.

>>>>

For a closed-shell reference determinant ($s_0=0$), hCI0 reduces to the restricted HF determinant.
The first non-trivial order is hCI1 (see left of Fig.~\ref{fig:hCI}), which accounts for all single excitations (as CIS)
plus all paired doubled excitations (two electrons promoted from the same occupied orbital into the same virtual orbital),
as shown in the top left of Fig.~\ref{fig:determinants}.
For both classes of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, hence the empty blocks at $s=0$ for odd $e$.
hCI1.5 augments the set of hCI1 determinants with the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 incorporates all classes of determinants where $\Ndet$ scale as $N^4$.
In total, it accounts for all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons,
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Likewise, in eCI, $\Ndet$ scales as $N^{2e}$.
The computational scaling along the two routes is shown in Table~\ref{tab:scaling}.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space specified by the reference determinant(s) and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the present eCI, sCI and FCI calculations.
In a given calculation, the energies are considered to be converged when the (largest) EN2 correction computed in the truncated Hilbert space
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delivering fairly converged energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we further evaluated the standard and the renormalized EN2 perturbative correction. \cite{Garniron_2019}
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for eCI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation.
The computational scaling associated with the CI+EN2 calculations is also presented in Table~\ref{tab:scaling}.

@@

We start the discussion with the excitations from closed-shell systems, with corresponding statistical errors shown in Table~\ref{tab:stat1}.
The first level of hCI, hCI1, produces poor excitation energies, with a MAE of \SI{1.07}{\eV}.
This model shares the same computational scaling with CIS, but includes the paired double excitations,
which clearly worsen the decent CIS results (MAE of \SI{0.61}{\eV}).
Moving one rank up, to hCI1.5, the MAE increases to \SI{1.91}{\eV}, and then reaches a maximum of \SI{3.53}{\eV} at the hCI2 level.
We notice that hCI2 delivers somewhat smaller errors than CISD (MAE of \SI{4.09}{\eV}), even though both are way too large.
From that point on, the next hCI models generate progressively better results.
Despite the improvement with respect to hCI2, hCI2.5 is still as inaccurate (MAE of \SI{1.95}{\eV}) as hCI1.5.
At the hCI3 level, significantly smaller errors are finally achieved (MAE of \SI{0.19}{\eV}).
This is close to that obtained with the eCI model of same order, CISDT (\SI{0.17}{\eV}).
Even smaller errors are produced at the hCI3.5 level (MAE of \SI{0.11}{\eV}), but at a considerable computational cost.

@@

We now shift to the discussion of excitations for open-shell systems.
The statistical errors are shown in Table~\ref{tab:stat2}.
The key difference on the calculations for open-shell excitations is that the same type of reference was employed for ground- and excited-state calculations, namely,
a single open-shell determinant, shown in the center panel of Fig.~\ref{fig:hCI}.
In contrast, the excited states from the closed-shell systems relied on different classes of reference determinants.
This accounts for the more accurate results observed for open-shell excitations and explains most of the differences with respect to excitations from closed-shell systems, as discussed in detail in the following.

@@

We refer back to Table~\ref{tab:stat1} to discuss the performance of sCI for excited states of closed-shell systems.
The first model, sCI2/sCI2, systematically overestimates the excitation energies, producing a MAE of \SI{1.34}{\eV}.
The same two factors discussed above for hCI explain such large errors.
First, the ground state description is favored due to the use of ground-state HF orbitals.
Indeed, with state-specific orbitals, the $\Delta$sCI2/sCI2 model reduces the MAE to \SI{0.79}{\eV}, though still overestimating the excitation energies.
Second, there is an unbalanced description of correlation for ground and excited states, associated with the classes of determinants.
While sCI2/sCI2 accounts for an additional $s=2$ sector for the ground state (which can be qualitatively described in the $s=0$ sector),
no addition pair of electrons is allowed to become unpaired in the excited state calculation, even though the state is qualitatively described by a determinant that is contained in the $s=2$ sector.
In other words, unpaired excitations are allowed to correlate the ground state, but not the excited state, given their respective closed-shell and open-shell characters,
thus creating a bias towards the former.

@@

The situation for open-shell systems is quite different, in close analogy to the previous discussion regarding hCI.
As shown in Table~\ref{tab:stat2}, sCI1 provides considerably more accurate excitation energies for open-shell than for closed-shell systems.
The reason should not be surprising at this point.
Both ground and excited states of open-shells can be qualitatively described by the same type of reference (a single $s=1$ open-shell determinant).
Moreover, the state-specific approach is superior.
sCI1 presents a MAE of \SI{0.39}{\eV} and overestimates the excitation energies (MSE of \SI{+0.24}{\eV}), in view of the bias introduced by the ground-state orbitals.
With state-specific orbitals, $\Delta$sCI1 reduces the MAE to \SI{0.27}{\eV} and the MSE to \SI{-0.11}{\eV}, which is smaller in absolute value than found for sCI1.
